---
title: "Capstone_Fitbit"
author: "Lakshmi Raghavan"
date: "2/2/2022"
output: html_document
---

```{r setup, include=FALSE}
install.packages("r package", repos = "http://cran.us.r-project.org")
install.packages('plyr', repos = "http://cran.us.r-project.org")

knitr::opts_chunk$set(echo = TRUE)

install.packages('tidyverse', repos = "http://cran.us.r-project.org")
install.packages('lubridate', repos = "http://cran.us.r-project.org")
install.packages('dplyr', repos = "http://cran.us.r-project.org")
install.packages('ggplot2', repos = "http://cran.us.r-project.org")
install.packages('tidyr', repos = "http://cran.us.r-project.org")
install.packages('kableExtra', repos = "http://cran.us.r-project.org")

library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
library(tidyr)
library(kableExtra)
```

## **Introduction**

#### The global fitness tracker market size is expected to reach USD 138.7 billion by 2028, registering a CAGR of 18.9% over the forecast period, according to a new report by Grand View Research, Inc. Some of the factors contributing to the market growth are the rising awareness regarding health & fitness, growing adoption of smartphones, availability of the internet, and increasing consumer disposable incomes.(prnewswire.com, August 2021)


## **Background**

#### This Kaggle data set contains personal fitness tracker from thirty fitbit users. Thirty eligible Fitbit users consented to the submission of personal tracker data, including minute-level output for physical activity, heart rate, and sleep monitoring. It includes information about daily activity, steps, and heart rate that can be used to explore users’ habits.(CC0: Public Domain, dataset made available through Mobius). This projects aims to identify some of the trends in fitness tracker usage using R programming Language. 


## **Importing Data**

```{r, echo=FALSE}

daily_activity <- read.csv("C:/Google Materials/Capstone/dataset/dailyActivity_merged.csv")
sleep_day <- read.csv("C:/Google Materials/Capstone/dataset/sleepDay_merged.csv")
weightlog <- read.csv("C:/Google Materials/Capstone/dataset/weightLogInfo_merged.csv")
heartrate <- read.csv("C:/Google Materials/Capstone/dataset/heartrate_seconds_merged.csv")

hourly_steps <- read.csv("C:/Google Materials/Capstone/dataset/hourlySteps_merged.csv")
hourly_intensities <- read.csv("C:/Google Materials/Capstone/dataset/hourlyIntensities_merged.csv")
hourly_calories <- read.csv("C:/Google Materials/Capstone/dataset/hourlyCalories_merged.csv")

hourly_one <- merge(hourly_steps, hourly_intensities, by=c ("Id", "ActivityHour"))

hourly_activity <- merge(hourly_one, hourly_calories, by=c ("Id", "ActivityHour"))

```
#### I have included 7 datasets from the original dataset, which contained 18 datasets overall. 

#### *First 4 datasets used in the orginal format*
  - dailyActivity_merged
  - sleepDay_merged
  - weightLogInfo_merged
  - heartrate_seconds_merged
  
#### *I have combined 3 hourly datasets into one hourly_activity dataset*
  - hourlySteps_merged
  - hourlyIntensities_merged
  - hourlyCalories_merged
  
## Distinct IDs and number of Rows

#### Before cleaning up the dataset, I wanted to see the dinstinct IDs and number of rows in each dataset. 

### daily_activity
```{r, echo=FALSE}
n_distinct(daily_activity$Id)
nrow(daily_activity)

```

### sleep_day
```{r,echo=FALSE}
n_distinct(sleep_day$Id)
nrow(sleep_day)

```

### WeightLogInfo
```{r, echo=FALSE}
n_distinct(weightlog$Id)
nrow(weightlog)

```

### heartrate
```{r, echo=FALSE}
n_distinct(heartrate$Id)
nrow(heartrate)


```

###hourly_activity
```{r, echo=FALSE}
n_distinct(hourly_activity$Id)
nrow(hourly_activity)
```

## Summary

### activity
```{r, echo=FALSE}
da_summary <- daily_activity %>%  
  select(TotalSteps,
         TotalDistance,TrackerDistance,
         SedentaryMinutes, Calories) %>%
  summary()
kable(da_summary, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

```

### Sleep
```{r, echo=FALSE}
sd_summary <- sleep_day %>%  
  select(TotalSleepRecords,
         TotalMinutesAsleep,TotalTimeInBed) %>%
  summary()
kable(sd_summary, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
```

### Weight log
```{r, echo=FALSE}
wl_summary <- weightlog %>%
  select(IsManualReport) %>%
  summary()
kable(wl_summary, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
```

### Heart rate
```{r, echo=FALSE}
hr_summary <- heartrate %>%
  select(Value) %>%
  summary()
kable(hr_summary, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
```

### Hourly activity
```{r, echo=FALSE}
ha_summary <- hourly_activity %>%
  select(StepTotal, TotalIntensity, AverageIntensity, Calories) %>%
  summary()
kable(ha_summary, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
```

## **Cleaning the Dataset**

#### Data integrity is the key to an efficient analysis. I started with identifying any duplicates present in the dataset and the N/A values.   

### Printing Duplicates

### *Daily_activity*
```{r, echo = FALSE}
sum(duplicated(daily_activity))
```

### *Sleep_day*
```{r, echo = FALSE}
sum(duplicated(sleep_day))
```

### *Weight Log
```{r, echo = FALSE}
sum(duplicated(weightlog))
```

#### Based on this, we see that the dataset, "Sleep_day" contains three duplicate rows of data, which need to be removed. 

### Removing Duplicates 

```{r, echo = FALSE}
sleep_day <- sleep_day %>%
  distinct() %>%
  drop_na()
```

### Confirming Removal of Duplicates 

```{r, echo = FALSE}
sum(duplicated(sleep_day))
```

### checking for NAs

```{r, echo = FALSE}
sum(is.na(daily_activity))
sum(is.na(sleep_day))
sum(is.na(weightlog))
sum(is.na(heartrate))
sum(is.na(hourly_activity))
```

### Formatting the Date to make sure its consistent everywhere

#### It was also observed that the Data column didn't have a consistent format, so I did the formattint to ensure consistency. 

```{r, echo = FALSE}
# dailyactivity
daily_activity$ActivityDate=as.POSIXct(daily_activity$ActivityDate, format="%m/%d/%Y", tz=Sys.timezone())
daily_activity$date <- format(daily_activity$ActivityDate, format = "%m/%d/%y")

# sleep
sleep_day$SleepDay=as.POSIXct(sleep_day$SleepDay, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
sleep_day$date <- format(sleep_day$SleepDay, format = "%m/%d/%y")

# weightlog
weightlog$Date=as.POSIXct(weightlog$Date, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
weightlog$time <- format(weightlog$Date, format = "%H:%M:%S")
weightlog$date <- format(weightlog$Date, format = "%m/%d/%y")

# heartrate
heartrate$Time=as.POSIXct(heartrate$Time, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
heartrate$time <- format(heartrate$Time, format = "%H:%M:%S")
heartrate$date <- format(heartrate$Time, format = "%m/%d/%y")

# HourlyActivity
hourly_activity$ActivityHour=as.POSIXct(hourly_activity$ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
hourly_activity$time <- format(hourly_activity$ActivityHour, format = "%H:%M:%S")
hourly_activity$date <- format(hourly_activity$ActivityHour, format = "%m/%d/%y")
```

### Verification if the date format is consistent

```{r, echo = FALSE}
# Verification if the date format is consistent
class(daily_activity$ActivityDate)
class(sleep_day$SleepDay)
class(weightlog$Date)
class(heartrate$Time)
class(hourly_activity$ActivityHour)
```

## *Analysis*

### Identifying Usage trends:

#### This chart illustrates the usage trends - how many days these tracker was used. 

```{r, echo=FALSE}
# Identifying Usage Trends 
daily_use_cat <- daily_activity %>%
  filter(TotalSteps >200 ) %>% 
  group_by(Id) %>%
  summarize(ActivityDate=sum(n())) %>%
  mutate(Usage = case_when(
    ActivityDate >= 1 & ActivityDate <= 14 ~ "Low Use",
    ActivityDate >= 15 & ActivityDate <= 21 ~ "Moderate Use", 
    ActivityDate >= 22 & ActivityDate <= 31 ~ "High Use")) %>% 
  mutate(Usage = factor(Usage, level = c('Low Use','Moderate Use','High Use'))) %>% 
  rename(daysused = ActivityDate) %>% 
  group_by(Usage)
tibble(head(daily_use_cat))

#Vizualising in a chart
#data transformation to create df that is readable to create a pie chart
##the 'scales' package is used here
daily_usage_trend <- daily_activity %>% 
  left_join(daily_use_cat, by = 'Id') %>%
  group_by(Usage) %>% 
  summarise(participants = n_distinct(Id)) %>% 
  mutate(perc = participants/sum(participants)) %>% 
  arrange(perc) %>% 
  mutate(perc = scales::percent(perc))
head(daily_usage_trend)


#Viz with a chart for distribution for different types of usage trends
ggplot(daily_usage_trend,aes(fill=Usage ,y = participants, x="")) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0)+
  scale_fill_brewer(palette='Accent')+
  theme_void()+
  theme(axis.title.x= element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5,vjust= -5, size = 20, face = "bold")) +
  geom_text(aes(label = perc, x=1.2),position = position_stack(vjust = 0.5))+
  labs(title="Usage Trend Distribution")+
  guides(fill = guide_legend(title = "Usage Type"))
  options(repr.plot.width = 1, repr.plot.height = 1)
```

### By day of the week(steps,sleep,calories burnt)

#### Next, I calculated the average steps, sleep and Calories burnt based on day of the week. 

### Steps
```{r, echo=FALSE}
dow_steps <- daily_activity %>%
  mutate(weekday = weekdays(ActivityDate))

dow_steps$weekday <-ordered(dow_steps$weekday, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday",
"Friday", "Saturday"))

dow_steps<-dow_steps%>%
  group_by(weekday) %>%
  summarize (daily_steps = mean(TotalSteps))

head(dow_steps)


ggplot(dow_steps) +
      geom_col(aes(weekday, daily_steps), fill = "#010101") +
      geom_hline(yintercept = 7500) +
      labs(title = "Average Daily steps per Day of the Week", x= "", y = "") +
      theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 1))
```

### Sleep

```{r, echo=FALSE}
dow_sleep <- sleep_day %>%
  mutate(weekday = weekdays(SleepDay))

dow_sleep$weekday <-ordered(dow_sleep$weekday, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday",
"Friday", "Saturday"))

dow_sleep <- dow_sleep %>%
  group_by(weekday) %>%
  summarize (daily_sleep = mean(TotalMinutesAsleep))

head(dow_sleep)

ggplot(dow_sleep) +
      geom_col(aes(weekday, daily_sleep), fill = "#010101") +
      geom_hline(yintercept = 600) +
      labs(title = "Average Minutes Asleep per Day of the Week", x= "", y = "") +
      theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 1))
```

### Calories

```{r, echo=FALSE}
dow_cal <- daily_activity %>%
  mutate(weekday = weekdays(ActivityDate))

dow_cal$weekday <- ordered(dow_cal$weekday, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday",
"Friday", "Saturday"))

dow_cal <- dow_cal %>%
  group_by(weekday) %>%
  summarize (daily_cal = mean(Calories))

head(dow_cal)


ggplot(dow_cal) +
      geom_col(aes(weekday, daily_cal), fill = "#010101") +
      geom_hline(yintercept = 3000) +
      labs(title = "Average Calories burnt per Day of the Week", x= "", y = "") +
      theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 1))
```

### Fitness Tracker usage throughout a day

#### Similarly, this chart shows the trend based on the number of winutes the fitness trackers were minutes worn by the consumer.

```{r, echo=FALSE}
usage_minutes <- daily_activity %>% 
  mutate(total_minutes = VeryActiveMinutes+FairlyActiveMinutes+LightlyActiveMinutes+SedentaryMinutes)%>%
  mutate (pctge_minutes_worn = (total_minutes/1440)*100) %>%
  mutate (worn = case_when(
    pctge_minutes_worn == 100 ~ "All day Usage",
    pctge_minutes_worn < 100 & pctge_minutes_worn >= 50~ "Moderate Usage", 
    pctge_minutes_worn < 50 & pctge_minutes_worn > 0 ~ "Less Usage"
  ))

um_summary <- head(usage_minutes)
kable(um_summary, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

usage_daily_minutes <- daily_activity %>% 
  left_join(usage_minutes, by = 'Id') %>%
  group_by(worn) %>% 
  summarise(participants = n_distinct(Id)) %>% 
  mutate(perc = participants/sum(participants)) %>% 
  arrange(perc) %>% 
  mutate(perc = scales::percent(perc))
udm_summary <- head(usage_daily_minutes)
kable(udm_summary, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

ggplot(usage_daily_minutes,aes(fill=worn ,y = participants, x="")) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0)+
  scale_fill_brewer(palette='Accent')+
  theme_void()+
  theme(axis.title.x= element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5,vjust= -5, size = 15, face = "bold")) +
  geom_text(aes(label = perc, x=1.2),position = position_stack(vjust = 0.5))+
  labs(title="Daily Usage Minutes Distribution")+
  guides(fill = guide_legend(title = "Usage Type"))
  options(repr.plot.width = 1, repr.plot.height = 1)
```


### Minutes Asleep Trend 

#### Likewise, trend in sleep patterns. 

```{r, echo=FALSE}
total_sleep <- sleep_day %>% 
  mutate(TotalMinAsleep = TotalMinutesAsleep) %>% 
  mutate (minutes_asleep = case_when(TotalMinutesAsleep <= 360 ~ "Not Enough sleep",
    TotalMinutesAsleep < 680 & TotalMinutesAsleep > 360 ~ "Moderate Sleep", 
    TotalMinutesAsleep >= 680 ~ "Enough Sleep"
  ))
  
ts_summary <- head(total_sleep)
kable(ts_summary, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

daily_sleep_trend <- sleep_day %>% 
  left_join(total_sleep, by = 'Id') %>%
  group_by(minutes_asleep) %>% 
  summarise(participants = n_distinct(Id)) %>% 
  mutate(perc = participants/sum(participants)) %>% 
  arrange(perc) %>% 
  mutate(perc = scales::percent(perc))
dst_summary <- head(daily_sleep_trend)
kable(dst_summary, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")

ggplot(daily_sleep_trend,aes(fill=minutes_asleep ,y = participants, x="")) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0)+
  scale_fill_brewer(palette='Accent')+
  theme_void()+
  theme(axis.title.x= element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5,vjust= -5, size = 15, face = "bold")) +
  geom_text(aes(label = perc, x=1.2),position = position_stack(vjust = 0.5))+
  labs(title="Average Minutes Asleep Distribution")+
  guides(fill = guide_legend(title = "Usage Type"))
  options(repr.plot.width = 1, repr.plot.height = 1)
```

### Correlation between variables. 

#### These next set of visualizations depicts the correlation between different set of variables. 
### Total Steps Vs Calories

```{r, echo=FALSE}
ggplot(usage_minutes, aes(x=TotalSteps, y=Calories))+
    geom_jitter() +
    geom_smooth(color = "green") + 
    labs(title = "Total Steps Vs Calories Burnt", x = "Total Steps", y= "Calories Burnt") +
      theme(panel.background = element_blank(),
      plot.title = element_text( size=14))
  
```

### Sedantary Minutes Vs Calories Burnt

```{r, echo=FALSE}
ggplot(usage_minutes, aes(x=SedentaryMinutes, y=Calories))+
    geom_jitter() +
    geom_smooth(color = "red") + 
    labs(title = "Sedantary Minutes Vs Calories Burnt", x = "Sedantary Minutes", y= "Calories Burnt") +
    theme(panel.background = element_blank(),
      plot.title = element_text( size=14))
```

### Minutes Worn Vs Calories Burnt 

```{r, echo=FALSE}
ggplot(usage_minutes, aes(x=total_minutes, y=Calories))+
    geom_jitter() +
    geom_smooth(color = "green") + 
    labs(title = "Minutes worn Vs Calories Burnt", x = "Minutes Worn", y= "Calories Burnt") +
    theme(panel.background = element_blank(),
        plot.title = element_text( size=14))
```

### Minutes worn Vs Logged Activities Distance 

```{r, echo=FALSE}
ggplot(usage_minutes, aes(x=total_minutes, y=LoggedActivitiesDistance))+
    geom_jitter() +
    geom_smooth(color = "red") + 
    labs(title = "Minutes worn Vs Logged Activities Distance", x = "Minutes Worn", y= "Logged Activities") +
    theme(panel.background = element_blank(),
        plot.title = element_text( size=14))
```


```{r, echo=FALSE}
##Merging Activity and Sleep
activity_sleep_data <- merge(sleep_day, daily_activity, by= c("Id", "date"))
```

### Correlation

#### Finally, this heatmap shows the correlation between the variables from the dataset daily_activity. 

```{r, echo=FALSE}
sleep_correlation <- activity_sleep_data[,c(5,6,8,9,10,11,16,17,18,19,20)]
sc_summary <- head(sleep_correlation)
kable(sc_summary, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
dim(sleep_correlation)
cor_matrix2 <- cor(sleep_correlation)
my_colors <- colorRampPalette(c("lightblue", "purple")) 
heatmap(cor_matrix2, Rowv = NA, Colv = NA, col = my_colors(100), margins = c(12, 12), cexRow=0.7,cexCol = 0.7)
#heatmap(cor_matrix2, display_numbers = T) 
#heatmap(data, col = my_colors(100)) 
#heatmap(data, Rowv = NA, Colv = NA) 
```

#### Note: The dataset was obtained from Kaggle, I have used R Studio Cloud for the entire Data Analysis and R Markdown for the Report Generation.
